---
title: "Columbia Payroll"
output:
  html_document:
    df_print: paged
---

```{r message=F}
library(tidyverse)
```

```{r}
payroll <- read_csv("data/como-payroll-2024.csv")
```

Get to know the data:

How many rows? 38,883\
What is one row? One paycheck\
What is in every column?

-   name (first and last)
-   department (actually supervisor's title)
-   job title
-   check number
-   check date
-   total pay (the other pay columns combined)
-   base, overtime, other, employer-paid (other pay categories)

What are the date boundaries of the data? (the city's fiscal year for 2024, Oct through Sept)

```{r}
range(payroll$check_date)
```

Questions about the data:

-   What is other pay? (a reporting question)

-   How to identify individual departments?

```{r}
payroll %>% count(department)
# since this is not a true department, it would take a lot of work to create a clean department column. 

# instead you can look for a particular department. for example, find all "department" values that have "police" or "pd":

# just police
payroll %>% 
  filter(grepl("police", department, ignore.case=T))

# police OR pd
payroll %>% 
  filter(grepl("police|pd", department, ignore.case=T))

# create a new table with just police:
police <- payroll %>% 
  filter(grepl("police|pd", department, ignore.case=T))
```

Questions for the data:

1\. Which job title gets paid the most non base pay? Which job title gets paid the most overtime?

```{r}
payroll %>% 
  group_by(job_title) %>% 
  summarise(non_base_pay = sum(overtime_pay+other_pay+employer_paid_benefits)) %>% 
  arrange(desc(non_base_pay))
# keep in mind that these aren't necessarily fair comparisons because we don't know how many people have each job title: the more employees, the higher the number is likely to be. 

# we could calculate mean(), but that is per paycheck, not per person, and so is not ideal: 
payroll %>% 
  group_by(job_title) %>% 
  summarise(non_base_pay = mean(overtime_pay+other_pay+employer_paid_benefits)) %>% 
  arrange(desc(non_base_pay))

```

It's better to create a dataset of people so that you can more accurately count people and find meaningful averages, medians etc.

```{r}
people <- payroll %>% 
# group by the variables that are consistent across paychecks
  group_by(last_name, first_name, department, job_title) %>% 
# summarise the variables that are specific to each paycheck (we can ignore check num and date)
  summarise(total_pay = sum(total_pay), 
            base_pay = sum(base_pay), 
            overtime_pay = sum(overtime_pay), 
            other_pay = sum(other_pay), 
            employer_paid_benefits = sum(employer_paid_benefits))

# always good to check for NAs in numbers columns because that will affect totals:
payroll %>% 
  filter(is.na(overtime_pay))
# no NAs

# now calculate average non base pay for people in each job title:
people %>% 
  group_by(job_title) %>% 
  summarise(non_base_pay = mean(overtime_pay+other_pay+employer_paid_benefits),
            employees = n()) %>% 
  arrange(desc(non_base_pay))
```

2.  How much is the city paying out in pensions every year?

```{r}
# how can we identify pension payments? Perhaps by using department, but would need to vet this with someone at the city. 
people %>% 
  filter(grepl("pension", department, ignore.case=T)) %>% 
  summarise(total = sum(total_pay))

# Note that we can't aggregate over the groups we created in the `people` variable unless we use ungroup():
people %>% 
  filter(grepl("pension", department, ignore.case=T)) %>% 
  ungroup() %>% 
  summarise(total = sum(total_pay))
```

3.  Which individual got paid the most?

```{r}
people %>% arrange(desc(total_pay))
```

4.  How many people are in different job titles? (though we really want to know about departments)

```{r}
people %>% 
  group_by(job_title) %>% 
  summarise(employees = n()) %>% 
  arrange(desc(employees))
```

5.  Which job title / department has the highest average salary?

6.  Who gets paid the least?
