---
title: "Hot 100 in class"
output: 
---

This dataset is based off of Billboard's weekly Hot 100 songs:
<https://www.billboard.com/charts/hot-100/>

It is compiled by Christian McDonald out of UT-Austin:
<https://github.com/utdata/rwd-billboard-data>

```{r message=F}
library(tidyverse)
```

```{r}
hot100 <- read_csv("data/hot-100-current.csv")
```

Do some assessment on the data. How many rows? What is one row? What's
in every column?

Rows: 347,300

One row: *one song* on the charts for *one week*.

**chart_week**: the week the chart was released (100 songs for each
week)

**current_week**: the ranking of that particular song for the chart_week

**title**: the title of the song

**performer**: who performs the song

**last_week**: the ranking of this particular song on last week's chart
(could be 0 or NA if it wasn't on the charts the week before)

**peak_pos**: the highest ranking that song has achieved as of
chart_week

**wks_on_chart**: how many weeks that song has been on the charts as of
chart_week.

ASSESSMENT 
```{r}
# What are the time boundaries of this data? 
hot100 %>% reframe(range(chart_week))

# How many weeks are in the data? 3,473 (or the total # of rows divided by 100)
hot100 %>% count(chart_week)

# You can also use distinct() to see unique values in any column: 
hot100 %>% distinct(chart_week)
```

distinct() is a tidyverse function so you can pipe data directly into it. It will give you a list of the unique values (if you specify a column) or a list of unique rows.

```{r}
# How many years are represented? Use the year() function to pull the year from the date:
hot100 %>% count(year(chart_week))

# using distinct():
hot100 %>% distinct(year(chart_week))

# you can also use distinct() with empty () to see unique rows (a good way to check for exact duplicates)
hot100 %>% distinct()

# further assessment: check to see if the values in current_week match what we expect (1 to 100):
hot100 %>% distinct(current_week)

# to look at performer, try sorting alphabetically:
hot100 %>% distinct(performer) %>% arrange(performer)

# since your results are limited to 1000 rows in the console below the code chunk, to see the full list you can create a new variable:
artists <- hot100 %>% distinct(performer) %>% arrange(performer)

view(artists)
```

n_distinct() is a base R summary function and therefore must be used inside summarise(). It will give you a count of unique values in a column. 

```{r}
# example: how many unique performers are there? 
hot100 %>% summarise(performers = n_distinct(performer))

# how many unique song titles are there for each performer? 
hot100 %>% 
  group_by(performer) %>% 
  summarise(unique_titles = n_distinct(title)) %>% 
  arrange(desc(unique_titles))

# note that some songs are rereleased with slightly different titles:
hot100 %>% 
  filter(performer == "Taylor Swift") %>% 
  distinct(title) %>% 
  arrange(title)
```
